---
title: "evo_ancestral_model_comp"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F)
```

# What is the ancestral state for blood-feeding? Does the blood-feeding trait have a directionality?
Hematophagy is the blood-feeding habit of various arthropods. Mosquitoes bloodfeed, but not all of them do so. Males never bloodfeed, most females bloodfeed to get nutrients to lay eggs. Interestingly, females do not always rely on blood, and some have even lost the capability to drink blood. 

When the females do not depend on blood, they develop capability to lay eggs without blood, a charactristic known as autogeny. Autogeny has been present in 

Whereas bloodfeeding as an ancestral state is well supported in literature, autogeny is not as clear. Some papers discuss it as a "seconary adaptation" (O'Meara), while others refer to it as "pleisiomorphic characteristic" (Coluzzi & Costantini).  Key papers summarizing concepts in autogeny discusses it as an "adaptive trait" (Downes).

In this paper, we explore the different degrees of hematophagy as defined by presence and absence of blood-feeding and autogeny. The goal of this paper is to further the understanding of evolution of bloodfeeding in its diffrent degrees. 

The published tree based on genetic and morphological characteristics (Reidnebach et al 2009) is used through these analysis:
```{r echo=FALSE}
### Import Tree & make ultrametric & plot tree===
library("phytools")
library(ggplot2)
cul.tree = read.nexus("Culicid_tree_2017.nex") #read nexus file
# turns the randomly generated branch lengths ultrametric
# cul.tree = chronopl(cul.tree, lambda=0)
tree<-cul.tree
plot(tree, cex=.7, no.margin=T)

cul.hema = read.csv('wings.hema.20171113.csv',sep=",")
cul.hema <- cul.hema[order(cul.hema$gs_sp),]
```
To test whether hematophagy is ancestral, I first followed Harmon's tutorial which assumes Equals rate of evolution. This may not a good model, I thus tested a few other models:

# Does character order improve model? Does it change if model assumes equal or different rates?

To start this question, let us categorize hematophagy as bloodfeeders (blood), facultative bloodfeeders (facbl), and non-bloodfeeders (nobld), and assuming blood-feeding is ancetral.

The models tested are Equal rates (ER), Symmetrical rates (SYM), and all-rates-different (ARD). The three models are inherently undordered, so I further coded all three models as ordered characters (i.e. a character can transition between A&B and B&C but not A and C).
```{r echo=FALSE}
  # http://blog.phytools.org/2016/03/comparing-alternative-models-for.html

# USE HEMA as 3 categories, assuming bloodfeeding is ancestral
x<-cul.hema$hema1
names(x)<-cul.hema$gs_sp
cols<-c("red","black","green3")
names(cols)<-c("blood","facbl","nobld")

# function to compare all models:
  ## Method 1: Estimate ancestral states under a Equal Rates (ER) model
get_all_aic<-function(x){
  fitER<-fitMk(tree,x,model="ER",type="discrete")
  # fitER
  fitSYM<-fitMk(tree,x,model="SYM",type="discrete")
  # fitSYM
  fitARD<-fitMk(tree,x,model="ARD",type="discrete")
  # fitARD

if (length(unique(x))==3) {
# ORDERED, REVERSIBLE
  ER.ordered<-matrix(c(0,1,0,1,0,1,0,1,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  # ER.ordered
  SYM.ordered<-matrix(c(0,1,0,1,0,2,0,2,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  # SYM.ordered
  ARD.ordered<-matrix(c(0,1,0,2,0,3,0,4,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  # ard.ordered
}

if (length(unique(x))==2) {
# ORDERED, REVERSIBLE
  ER.ordered<-matrix(c(0,0,1,0),2,2,dimnames=list(c("blood","nobld"),c("blood","nobld")))
  # ER.ordered
  SYM.ordered<-matrix(c(0,0,1,0),2,2,dimnames=list(c("blood","nobld"),c("blood","nobld")))
  # SYM.ordered
  ARD.ordered<-matrix(c(0,0,1,0),2,2,dimnames=list(c("blood","nobld"),c("blood","nobld")))
  # ard.ordered
}

fitER_ord<-fitMk(tree,x,model=ER.ordered,type="discrete")
# fitER_ord
fitSYM_ord<-fitMk(tree,x,model=SYM.ordered,type="discrete")
# fitSYM_ord
fitARD_ord<-fitMk(tree,x,model=ARD.ordered,type="discrete")
# fitARD_ord

aic<-as.matrix(sapply(list(ER=fitER,SYM=fitSYM,ARD=fitARD,
                           ER.ord=fitER_ord,SYM.ord=fitSYM_ord,ARD.ord=fitARD_ord
                           ),AIC))
names(aic)<-rownames(aic)
sort(aic)
}
get_all_aic(x)

```

Above is the sorted AIC scores of each model. This suggests that 1) models that assumed order were better fit than unordered models, and 2) all-rates-diferent models were best among the ordered models. 

## What if character is irreversible? Does it change if the model assumes equal or different rates?
We further test if irreversibility of the state changes would improve the model.
```{r}
# ORDERED, IRREVERSIBLE
  #BL->FL->NB
  irrBL2NB<-matrix(c(0,0,0,1,0,0,0,1,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  #NB->FL->BL #expect AIC to be worst this one
  irrNB2BL<-matrix(c(0,1,0,0,0,1,0,0,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  #FL->BL OR FL->NB
  irrFB2BLNB<-matrix(c(0,1,0,0,0,0,0,1,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  
fitirrBL2NB<-fitMk(tree,x,model=irrBL2NB,type="discrete")
fitirrNB2BL<-fitMk(tree,x,model=irrNB2BL,type="discrete")
fitirrFB2BLNB<-fitMk(tree,x,model=irrFB2BLNB,type="discrete")

## Does changing the irreversible rates to unequal rates change the outcome?  
  irrBL2NB2<-matrix(c(0,0,0,1,0,0,0,2,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  irrNB2BL2<-matrix(c(0,1,0,0,0,2,0,0,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  #FL->BL OR FL->NB
  irrFB2BLNB2<-matrix(c(0,1,0,0,0,0,0,2,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
fitirrBL2NB2<-fitMk(tree,x,model=irrBL2NB2,type="discrete")
fitirrNB2BL2<-fitMk(tree,x,model=irrNB2BL2,type="discrete")
fitirrFB2BLNB2<-fitMk(tree,x,model=irrFB2BLNB2,type="discrete")
# AIC(fitirrBL2NB2) #ERODES IT
# AIC(fitirrNB2BL2) #IMPROVES IT
# AIC(fitirrFB2BLNB2) # IMPROVES IT

aic<-as.matrix(sapply(list(BL2NB.irrER=fitirrBL2NB, 
                           NB2BL.irrER=fitirrNB2BL, 
                           FB2BLNB.irrER=fitirrFB2BLNB,
                           BL2NB.irrADR=fitirrBL2NB2, 
                           NB2BL.irrADR=fitirrNB2BL2, 
                           FB2BLNB.irrADR=fitirrFB2BLNB2
                           ),AIC))
names(aic)<-rownames(aic)
sort(aic)
```

Regardless of the assumption about the rate of evolution (ER or ADR), including irreversibility in the model worsens the model. 

Thus, for subsequent analysis, we shall make use of the model that assumes all-rates-different for evolutionary rates, and order (or directionality) in the model. The Qmatrix shall be:
```{r}
ARD.ordered
```

## What if hema character has 2 states?
```{r}
x<-cul.hema$hema3
names(x)<-cul.hema$gs_sp
get_all_aic(x)
```
Results seem to imply that it does not really matter, all models for the case where hema has 2 states may be perform similarly. An ordered binary state is just a 2x2 matrix with a single change, thus the "ordered" models were all the same regardless of assumption of rate of change.

# Now what if we just observe the consequences of autogeny, independent of bloodfeeding?
The issue with the original hematophagy states is that it relies on two descriptions in literature: 1) whether the mosquito is attracted to a host or seen imbibing on blood, and 2) whether the mosquito is autogenous, aka capable of laying eggs without blood. 

So let's trim the tree to a smaller tree, removing species without Autogeny data.
```{r}

drop<-as.character(cul.hema$gs_sp[cul.hema$hema4=="blood"])
tree.drop<-drop.tip(tree,drop)
# dev.new()
# par(mfrow=c(2,1))
# plotTree(tree)
plotTree(tree.drop)
tree<-tree.drop

x<-cul.hema$hema1[!cul.hema$hema4=="blood"]
names(x)<-cul.hema$gs_sp[!cul.hema$hema4=="blood"]
cols<-c("red","black","green3")
names(cols)<-c("blood","facbl","nobld")

get_all_aic(x)
```

