---
title: "Ancestral state reconstruction of hematophagy"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F)
```
# Newtitle4
untitled

# What is the ancestral state for blood-feeding?
Hematophagy is the blood-feeding habit of various arthropods. Mosquitoes bloodfeed, but not all of them do so. Males never bloodfeed, most females bloodfeed to get nutrients to lay eggs. Interestingly, females do not always rely on blood, and some have even lost the capability to drink blood. 

When the females do not depend on blood, they develop capability to lay eggs without blood, a charactristic known as autogeny. Autogeny has been present in 

Whereas bloodfeeding as an ancestral state is well supported in literature, autogeny is not as clear. Some papers discuss it as a "seconary adaptation" (O'Meara), while others refer to it as "pleisiomorphic characteristic" (Coluzzi & Costantini).  Key papers summarizing concepts in autogeny discusses it as an "adaptive trait" (Downes).

In this paper, we explore the different degrees of hematophagy as defined by presence and absence of blood-feeding and autogeny. The goal of this paper is to further the understanding of evolution of bloodfeeding in its diffrent degrees. 

The published tree based on genetic and morphological characteristics (Reidnebach et al 2009) is used through these analysis:
```{r echo=FALSE}
### Import Tree & make ultrametric & plot tree===
library("phytools")
library(ggplot2)
cul.tree = read.nexus("Culicid_tree_2017.nex") #read nexus file
# turns the randomly generated branch lengths ultrametric
cul.tree = chronopl(cul.tree, lambda=0)
tree<-cul.tree
plot(tree, cex=.7, no.margin=T)

cul.hema = read.csv('wings.hema.20171113.csv',sep=",")
cul.hema <- cul.hema[order(cul.hema$gs_sp),]

## WRITE FUNCTION to plot ancestral state ==== 
plot_ancestral_rec<-function(x,cols){
  # FOLLOW THESE INSTRUCTIONS
  # http://www.phytools.org/eqg2015/asr.html
  # dev.new()
  par(mfrow=c(2,1))
  # ANCESTRAL STATE RECONSTRUCTTION

  ## Method 1: Estimate ancestral states under a Equal Rates (ER) model
  # Let's fit a single-rate model & reconstruct ancestral states at internal nodes in the tree.
  ARD.ordered<-matrix(c(0,1,0,2,0,3,0,4,0),3,3,dimnames=list(c("blood","facbl","nobld"),c("blood","facbl","nobld")))
  fit<-ace(x,tree,model=ARD.ordered,type="discrete")
  # fit<-fastAnc(tree,x,vars=TRUE,CI=TRUE)
  fit
  round(fit$lik.anc,3)
  # The element lik.anc gives us the marginal ancestral states, also known as the 'empirical Bayesian posterior probabilities.'
  # It is fairly straightforward to overlay these posterior probabilities on the tree:
  plotTree(tree,fsize=0.8,ftype="i")
  nodelabels(node=1:tree$Nnode+Ntip(tree),
             pie=fit$lik.anc,piecol=cols,cex=0.5)
  tiplabels(pie=to.matrix(x,sort(unique(x))),piecol=cols,cex=0.3)
  add.simmap.legend(colors=cols,prompt=FALSE,x=0.9*par()$usr[1], 
                    y=-max(nodeHeights(tree))+8,fsize=0.7)
  
  
  ## Method 2: MCMC approach/ Stochastic character mapping
  # An alternative approach to the one outline above is to use an MCMC approach to sample character histories from their posterior probability distribution. This is called stochastic character mapping (Huelsenbeck et al. 2003). The model is the same but in this case we get a sample of unambiguous histories for our discrete character's evolution on the tree - rather than a probability distribution for the character at nodes.
  
  # Simulate single stochastic character map using empirical Bayes method
  # mtree<-make.simmap(tree,x,model="ER")
  # mtree
  # plot(mtree,cols,fsize=0.8,ftype="i")
  # add.simmap.legend(colors=cols,prompt=FALSE,x=0.9*par()$usr[1],
  #                   y=-max(nodeHeights(tree)),fsize=0.8)
  
  #### SIMULATIONSSSSS using baesyian mdels:
  # A single stochastic character map does not mean a whole lot in isolation - we need to look at the whole distribution from a sample of stochastic maps. This can be a bit overwhelming. The following code generates 100 stochastic character maps from our dataset, which could be plotted but I removed the actual plot and saved the simulation:
  ## make.simmap is sampling character histories conditioned on the transition matrix
  mtrees<-make.simmap(tree,x,model=ARD.ordered,nsim=100,pi="estimated")
  ## plot all simulated trees
  # par(mfrow=c(10,10))
  # null<-sapply(mtrees,plot,colors=cols,lwd=1,ftype="off")
  
  # It's possible to summarize a set of stochastic maps in a much more meaningful way. For instance, we can estimate the number of changes of each type, the proportion of time spent in each state, and the posterior probabilities that each internal node is in each state, under our model. For example:
  pd<-summary(mtrees,plot=FALSE)
  # pd 
  # plot(pd,fsize=0.7,ftype="i")
  
  ## now let's plot a random map, and overlay the posterior probabilities
  plot(mtrees[[1]],cols,fsize=0.8,ftype="i")
  
  nodelabels(pie=pd$ace,piecol=cols,cex=0.5)
  add.simmap.legend(colors=cols,prompt=FALSE,x=0.9*par()$usr[1],
                    y=-max(nodeHeights(tree))+8,fsize=0.8)
  
  # plot(fit$lik.anc,pd$ace,xlab="marginal ancestral states",
  #      ylab="posterior probabilities from stochastic mapping")
  # # lines(c(0,1),c(0,1),lty="dashed",col="red",lwd=2)
  
  ## Finally, since we obtained these inferences under exactly the same model, let's compare the posterior probabilities from stochastic mapping with our marginal ancestral states. In the former case, our probabilities were obtained by sampling from the joint (rather than marginal) probability distribution for the ancestral states.
  # plot(fit$lik.anc,pd$ace,xlab="marginal ancestral states",
       # ylab="posterior probabilities from stochastic mapping")
  # lines(c(0,1),c(0,1),lty="dashed",col="red",lwd=2)
}

``` 

## Ancestral state of hematophagy as 3 categories, assuming bloodfeeding is ancestral

```{r , fig.width=5, fig.height=5.75}
## hema1:	3 categories, NA= blood
x<-cul.hema$hema1
names(x)<-cul.hema$gs_sp
cols<-c("red","black","green3")
names(cols)<-c("blood","facbl","nobld")
plot_ancestral_rec(x,cols)
```

## Ancestral state of hematophagy as 3 categories, assuming facultative bloodfeeding is ancestral
* current code is having some issues, but the output should be similar to the previous state, +/- 0.3 of differece between blood-feeders and facultative blood-feeders. Update TBD.
```{r , fig.width=5, fig.height=5.75,eval = F }
# hema2:	3 categories, NA= facbl
# x<-cul.hema$hema2
# names(x)<-cul.hema$gs_sp
# cols<-c("red","black","green3")
# names(cols)<-c("blood","facbl","nobld")
# plot_ancestral_rec(x,cols)
```

## Ancestral state of hematophagy as 2 categories
```{r , fig.width=5, fig.height=5.75}
# hema3:	2 categories
x<-cul.hema$hema3
names(x)<-cul.hema$gs_sp
cols<-c("red","green3")
names(cols)<-c("blood","nobld")
plot_ancestral_rec(x,cols)
```

## Ancestral state of hematophagy as 4 categories

```{r , fig.width=5, fig.height=5.75}
# hema4:	4 categories
x<-cul.hema$hema4
names(x)<-cul.hema$gs_sp
cols<-c("red","black","green3","magenta")
names(cols)<-c("blood","facbl","nobld","oblig")
plot_ancestral_rec(x,cols)
```

# What if we trim the tree to look at autogeny?

```{r}
# [!cul.hema$hema4=="blood"]
drop<-as.character(cul.hema$gs_sp[cul.hema$hema4=="blood"])
tree<-drop.tip(tree,drop)

x<-cul.hema$hema1[!cul.hema$hema4=="blood"]
names(x)<-cul.hema$gs_sp[!cul.hema$hema4=="blood"]
cols<-c("red","black","green3")
names(cols)<-c("blood","facbl","nobld")
plot_ancestral_rec(x,cols)
```


